---
Title: "bact_differential_OTUs"
Author: Henry Paz (henry.paz@huskers.unl.edu)
Output:
  html_document:
  keep_md: yes
---

Generate bacteria relative abundance OTU table.

```{r, engine='bash'}
#convert from biom to txt
biom convert -i biom_files/bact_otutable_final_rarefied.biom --to-tsv --header-key taxonomy -o biom_files/bact_otutable_final_rarefied.txt
```  

```{r}
#load packages
library(tidyverse)

##create data set 
bact_otutable <- read_tsv("biom_files/bact_otutable_final_rarefied.txt", skip=1)

bact_otuid_tax <- bact_otutable[,c(1,232)]
bact_samples <- bact_otutable[,-c(1,232)]

#relative abundance
bact_relabundance <- bact_samples %>%
  mutate_all(funs(. / sum(.)))

bact_relabun_otutable <- merge(bact_otuid_tax, bact_relabundance, by="row.names", all=T) 
bact_relabun_otutable <- bact_relabun_otutable[,-1]

#make taxonomy last column
bact_relabun_otutable <- bact_relabun_otutable %>%
  select(-taxonomy, taxonomy)

#generate file
write_tsv(bact_relabun_otutable, "biom_files/bact_relabun_otutable.txt", col_names=T)
```

Generate bacteria discovery population.

```{r, engine='bash'}
#convert from biom to txt
biom convert -i biom_files/bact_relabun_otutable.txt --to-json --table-type="OTU table" --process-obs-metadata taxonomy -o biom_files/bact_relabun_otutable.biom

#filter core OTUs
filter_otus_from_otu_table.py -i biom_files/bact_relabun_otutable.biom --otu_ids_to_exclude_fp filter/bact_core_filter.txt --negate_ids_to_exclude -o biom_files/bact_relabun_core.biom

#create discovery data set
filter_samples_from_otu_table.py -i biom_files/bact_relabun_core.biom --sample_id_fp filter/total_extremes.txt -o biom_files/bact_relabun_discovery.biom

#Convert from biom to txt format 
biom convert -i biom_files/bact_relabun_discovery.biom --to-tsv --header-key taxonomy -o biom_files/bact_relabun_discovery.txt

sed 's/#OTU ID/OTUID/g' biom_files/bact_relabun_discovery.txt > r_inputs/bact_relabun_discovery.txt
``` 

Generate LEfSe input file.

```{r}
library(tidyverse)

##create data set bactera mapping 
bact_mapping <- read_tsv("r_inputs/bact_mapping.txt") %>% 
  select(SampleID, TrtForageQuality)

##create data set bacteria extremes
bact_extremes <- read_tsv("filter/total_extremes.txt", col_names = ("SampleID")) 

#merge data sets 
bact_trt <- left_join(bact_extremes, bact_mapping, by = "SampleID") 

#remove samples 
bact_trt <- bact_trt %>%
  filter(SampleID != "8791.8.d63.20NormalMDGS" & SampleID != "8847.2.d63.HQ40MDGSNoRum" & SampleID != "8997.8.d63.20NormalMDGS")

#transpose data set
bact_trt_trans <- bact_trt %>%
  t %>%
  as.data.frame(stringsAsFactors=F)

#set column names
colnames(bact_trt_trans) = bact_trt_trans[1, ]
bact_trt_trans = bact_trt_trans[-1, ]

#add OTUID and taxonomy columns to match relative abundance column headers
bact_trt_trans <- bact_trt_trans %>%
  mutate(OTUID = "ForageQuality",
         taxonomy = "") 

##create data set bacteria discovery set
bact_discovery <- read_tsv("r_inputs/bact_relabun_discovery.txt", skip=1) 

#merge data sets 
bact_lefse <- rbind(bact_trt_trans, bact_discovery) 

#lefse format
bact_lefse <- bact_lefse %>% 
  select(-taxonomy) %>% 
  select(OTUID, everything()) 

bact_lefse[nrow(bact_lefse) + 1, ] <- colnames(bact_lefse)

bact_lefse <- bact_lefse[c(1,574,2:573),]

#generate file
write_tsv(bact_lefse, "lefse/bact_lefse_input.txt", col_names = F)
```

Identify bacterial differentially abundant OTUs. Results from LEfSe are provided in the Github [repository](https://github.com/enriquepaz/2018_methane_steers) within the lefse directory and were generated as described below.

```{r, engine='bash', eval=FALSE}
#convert the input data to LEfSe format
lefse-format_input.py lefse/bact_lefse_input.txt lefse/bact_lefse_inputin.txt -c 1 -u 2 -o 1000000

#performs LEfSe statistical analyses
run_lefse.py lefse/bact_lefse_inputin.txt lefse/bact_lefse_output.txt -a 1.0 -w 1.0 -l 0.0 --wilc 1 -y 1
```

Subset significant differentially abundant OTUs.

```{r}
#load packages
library(tidyverse)

##create data set 
bact_lefse <- read_tsv("lefse/bact_lefse_output.txt", col_names=c("OTUid", "loghighestclassaverage", "class", "LDA", "Pvalue"))

bact_lefse$OTUid <- gsub("f_","", bact_lefse$OTUid)

#adjust p-values using the Benjamini & Hochberg alias false discivery rate
bact_lefse <- bact_lefse %>%
  arrange (Pvalue) %>%
  mutate (fdr = p.adjust(Pvalue, method = "fdr"))

#filter significant differentially abundant OTUs
bact_differential <- bact_lefse %>%
  na.omit() %>% 
  filter (LDA >=2 & fdr <= 0.05 ) %>%
  arrange (desc(LDA)) %>% 
  slice(1:80) %>% 
  select (OTUid)

#generate file
write_tsv(bact_differential, "lefse/bact_differential_otus.txt", col_names = F)
```

Generate heatmaps.

```{r, engine='bash'}
##create data set 
filter_otus_from_otu_table.py -i biom_files/bact_relabun_discovery.biom -e lefse/bact_differential_otus.txt --negate_ids_to_exclude -o biom_files/bact_differential_otus_table.biom

#Convert from biom to txt format 
biom convert -i biom_files/bact_differential_otus_table.biom --to-tsv --header-key taxonomy -o biom_files/bact_differential_otus_table.txt
```

```{r}
library(tidyverse)
library(gplots)
library(Heatplus)
library(vegan)
library(RColorBrewer)

#generate heatmap for all data bacteria
##create data set differential otus
bact_diff_otus <- read_tsv("biom_files/bact_differential_otus_table.txt", skip=1) 

bact_genus <- bact_diff_otus %>%
  rename(OTUID = "#OTU ID") %>% 
  separate(taxonomy,
           into = c("kingdom","phylum","class","order","family","genus","specie"),
           sep=";") %>%
  select(-c(kingdom,phylum,class,order,family,specie))

#format genus column data
bact_genus$genus <- gsub("^\\s","", bact_genus$genus)
bact_genus$genus <- gsub("g__","", bact_genus$genus)
bact_genus$genus <- gsub("\\[","", bact_genus$genus)
bact_genus$genus <- gsub("\\]","", bact_genus$genus)
bact_genus$genus <- gsub("^k.*","", bact_genus$genus) 
bact_genus$genus <- gsub("^p.*","", bact_genus$genus) 
bact_genus$genus <- gsub("^c.*","", bact_genus$genus) 
bact_genus$genus <- gsub("^o.*","", bact_genus$genus) 
bact_genus$genus <- gsub("^f.*","", bact_genus$genus) 
bact_genus$genus <- gsub("^$","Unclassified", bact_genus$genus)

#format to data frame
bact_genus_df <- as.data.frame(bact_genus)
row.names(bact_genus_df) <- bact_genus_df$OTUID
bact_genus_df <- bact_genus_df[, -1]

#obtain treatment metadata from mapping file
##create data set bacteria mapping file
bact_mapping <- read_tsv("r_inputs/bact_mapping.txt") %>% 
  select(SampleID, TrtForageQuality)

##create data set bacteria extremes
bact_extremes <- read_tsv("filter/total_extremes.txt", col_names = ("SampleID")) 

#merge data sets 
bact_trt <- left_join(bact_extremes, bact_mapping, by = "SampleID") 

#remove samples 
bact_trt <- bact_trt %>%
  filter(SampleID != "8791.8.d63.20NormalMDGS" & SampleID != "8847.2.d63.HQ40MDGSNoRum" & SampleID != "8997.8.d63.20NormalMDGS")

#transpose data set
bact_trt_trans <- bact_trt %>%
  t %>%
  as.data.frame(stringsAsFactors=F)

#set column names
colnames(bact_trt_trans) = bact_trt_trans[1, ]
bact_trt_trans = bact_trt_trans[-1, ]

#add OTUID and genus columns to match relative abundance column headers
bact_trt_trans <- bact_trt_trans %>%
  mutate(genus = "genus") 

#merge differentila OTUs and treatment metadata sets 
bact_genus_trt <- rbind(bact_trt_trans, bact_genus_df) 
trtnames <- bact_genus_trt[1,1:63]
bact_genus_trt <- bact_genus_trt[-1,]
bact_taxa <- subset(bact_genus_trt, select=c(genus))
bact_samples = bact_genus_trt[,-64]
colnames(bact_samples) <- trtnames
bact_samples_trans <- as.data.frame(t(bact_samples), stringsAsFactors=F)

#convert data frame to numeric
bact_samples_trans[, 1:80] <- sapply(bact_samples_trans[, 1:80], as.numeric)

#add a dendrogram for the samples
#calculate the Bray-Curtis dissimilarity matrix on the full dataset
data.dist <- vegdist(bact_samples_trans, method = "bray")

#average linkage hierarchical clustering
row.clus <- hclust(data.dist, "aver")

#add a dendrogram for the taxa
data.dist.g <- vegdist(t(bact_samples_trans), method = "bray")
col.clus <- hclust(data.dist.g, "aver")

#color palette
scalewhiteblue <- colorRampPalette(c("white", "blue"), space = "rgb")(100)

#generate figure
pdf("figures/figure10a.pdf", height=8, width=9)
heatmap.2(as.matrix(bact_samples_trans), Rowv = as.dendrogram(row.clus), Colv = as.dendrogram(col.clus), labCol="", xlab="OTUs", ylab="Forage Quality", col = scalewhiteblue, trace="none", density.info="none", margins = c(2, 6), lhei = c(2, 10))
dev.off()

#generate heatmap for subset data bacteria
#determine the maximum relative abundance for each column
maxab <- apply(bact_samples_trans, 2, max)

#remove the genera with less than 1% as their maximum relative abundance
n1 <- names(which(maxab < 0.01))
bact_samples_filtered <- bact_samples_trans[, -which(names(bact_samples_trans) %in% n1)]

#add a dendrogram for the samples
#calculate the Bray-Curtis dissimilarity matrix on the full dataset
data.dist_sub <- vegdist(bact_samples_filtered, method = "bray")

#average linkage hierarchical clustering
row.clus_sub <- hclust(data.dist_sub, "aver")

#add a dendrogram for the taxa
data.dist.g_sub <- vegdist(t(bact_samples_filtered), method = "bray")
col.clus_sub <- hclust(data.dist.g_sub, "aver")

#generate figure
pdf("figures/figure10b.pdf", height=8, width=9)
heatmap.2(as.matrix(bact_samples_filtered), Rowv = as.dendrogram(row.clus_sub), Colv = as.dendrogram(col.clus_sub), labCol=bact_taxa$genus, xlab="OTUs", ylab="Forage Quality", col = scalewhiteblue, trace="none", density.info="none", margins = c(7.5, 6), lhei = c(2, 10))
dev.off()
```
